---
title: "Update of package openSTARS"
author: "Mira Kattwinkel"
date: "May 10, 2018"
output: html_document
bibliography: bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## openSTARS prepares GIS data for stream network analysis
'openSTARS' is based on open source software ('R' and 'GRASS GIS') to prepare the
input data for the 'R' package 'SSN' for Spatial Statistical Modeling on Stream Networks.
It is intended as an alternative to the 'ArcGIS' toolbox 'STARS' [@peterson2014]. 

'SSN' [@hoef2014, [link](http://www.fs.fed.us/rm/boise/AWAE/projects/SpatialStreamNetworks.shtml)]
enables the statistical analysis of stream monitoring data based on regression models
and accounts for patterns of spatial autocorrelation among sampling locations.
'openSTARS' formates the required GIS data to be read into 'SSN'.

```{r prepare_plot, echo = FALSE, eval = TRUE, message=FALSE, results='hide'}
library(openSTARS)
if(.Platform$OS.type == "windows"){
  gisbase = "c:/Program Files/GRASS GIS 7.4.0"
} else {
  gisbase = "/usr/lib/grass74/"
}
initGRASS(gisBase = gisbase,
          home = tempdir(),
          override = TRUE)

dem_path <- system.file("extdata", "nc", "elev_ned_30m.tif", package = "openSTARS")
sites_path <- system.file("extdata", "nc", "sites_nc.shp", package = "openSTARS")

setup_grass_environment(dem = dem_path)
import_data(dem = dem_path, sites = sites_path)
derive_streams()
calc_edges()
calc_sites()
calc_prediction_sites(predictions = "preds", nsites = 100, netIDs = 15 )

dem <- readRAST("dem", ignore.stderr = TRUE)
sites <- readVECT("sites", ignore.stderr = TRUE)
edges <- readVECT("edges", ignore.stderr = TRUE)
pred_sites <- readVECT("preds", ignore.stderr = TRUE)
```

<div style = "float:right; position: relative;"> 
``` {r do_plot, echo = FALSE, eval = TRUE, fig.height = 4.2, fig.width = 4.2, fig.align = "right"}
plot(dem, col = terrain.colors(20), what = "image")
lines(edges, col = "blue")
points(sites, pch = 19, col = "red")
points(pred_sites, pch = 20, bg = "black")
legend(x = par("usr")[1]*1.0008, y = par("usr")[3]*1.018, col = c(2,1,"blue"), pch = c(19,20,NA), lwd=c(NA,NA,2),
       legend = c("sampling sites", "created prediction sites","derived streams"), cex = 0.85)
```
</div>

## Updates
* Include vector data as potential predictors in `calc_attributes_edges` and calc_attributes_sites_exact`
* Some changes in `import_data`to ensure the correct projection and extend of all data
* Several bug fixes

## Input data
Required:

* sampling locations (vector data, e.g., ESRI shapefile)
* digital elevation model (raster data, e.g., tiff)

Optional:

* stream network (vector data, e.g., ESRI shapefile)
* predictions sites (vector data, e.g., ESRI shapefile)
* potential explanatory variables (raster or vector data, e.g., tiff)

## Workflow
A full example can be found [here](https://github.com/MiKatt/openSTARS/blob/master/README.md).

## References